---
title: "Analyzing a small RNA-Seq dataset with CLAMP"
author: "Maria Chikina"
package: CLAMP
output: 
  BiocStyle::html_document:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Analyzing a small RNA-Seq dataset with CLAMP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache=FALSE)
library(CLAMP)
```

# Overview

This vignette provides a small-scale example demonstrating the main workflow of the **CLAMP** package  
(**C**ompressed **L**atent-variable **A**pproach for **M**assive data **P**rocessing).  
CLAMP is designed to extract interpretable latent variables from high-dimensional transcriptomic data by combining:

1. **Unsupervised matrix factorization** (`CLAMPbase`) for dimensionality reduction, and  
2. **Pathway-guided refinement** (`CLAMPfull`) that integrates biological prior knowledge to improve interpretability.

In this example, we analyze a small human whole blood RNA-Seq dataset provided within the package.  
We illustrate how to:

- Load and normalize the data,  
- Incorporate prior knowledge from public databases (e.g., KEGG, CellMarker),  
- Fit both the baseline and extended CLAMP models, and  
- Compare the results across implementations, including memory-efficient **Filebacked Big Matrix (FBM)** input.


## 1. Load Data and Normalize

```{r}
data("dataWholeBlood")
data("majorCellTypes")
data('celltypeTargets')

# Scale each gene to mean 0 and variance 1
dataWholeBlood <- tscale(dataWholeBlood)
```

## 2. Load Prior Knowledge Matrices

```{r}
# Download pathway and cell marker libraries from Enrichr
gmtList <- list(
  CellMarkers = getGMT(
    "https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=CellMarker_2024",
    "CellMarker_2024"
  ),
  KEGG = getGMT(
    "https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=KEGG_2021_Human",
    "KEGG_2021_Human"
  )
)

# Combine into a single sparse matrix
pathMatCell <- gmtListToSparseMat(gmtList)

# Load additional xCell reference matrix
data("xCell")

# Match pathways to the gene space of whole blood
matchedPathsWB <- getMatchedPathwayMatList(
  pathMatCell, xCell,
  new.genes = rownames(dataWholeBlood),
  min.genes = 2
)
```

## 3. Initialize Model with SVD

```{r}
# Compute truncated SVD for initialization
set.seed(1);dataWholeBlood.svd=rsvd::rsvd(dataWholeBlood)
```

## 4. Fit Baseline and Extended CLAMP Models

```{r}
# Fit baseline CLAMP
suppressMessages({
  wb.clamp.base <- CLAMPbase(
    dataWholeBlood,
    svdres = dataWholeBlood.svd,
    trace = FALSE,
    adaptive.p = 0.05
  )
})
```

```{r}
# Fit variance-prior extended CLAMP model
suppressMessages({
  wb.clamp.full.vp <- CLAMPfull(
    dataWholeBlood,
    priorMat = matchedPathsWB,
    clamp.base.result = wb.clamp.base,
    trace = TRUE, use_cpp = TRUE
  )
})
```

## 5. Compare Models

This plot compares the maximum Spearman correlation for each major blood cell type between the baseline (`CLAMPbase`) and the prior-guided (`CLAMPfull`) models.  
Points above the red dashed line indicate improved correspondence when biological priors are included.  
Most cell types show higher correlations under `CLAMPfull`, suggesting that integrating pathway information helps capture more biologically meaningful latent variables.

```{r fig.width=6, fig.height=5}
output <- compareBs(
  wb.clamp.base,
  wb.clamp.full.vp,
  celltypeTargets,
  "CLAMPbase",
  "CLAMPfull",
  m = "s"
)

output$plot
```

## 6. Model Comparison Summary

```{r}
knitr::kable(output$df, format = "markdown")
```

## 7. Test FBM functionality

```{r}
dataWholeBloodFBM=bigstatsr::as_FBM(dataWholeBlood)
suppressMessages({
  wb.clamp.full.fbm <- CLAMPfull(
    dataWholeBloodFBM,
    priorMat = matchedPathsWB,
    clamp.base.result = wb.clamp.base,
    trace = TRUE, use_cpp = TRUE
  )
})

```

Results are the same!

```{r fig.width=6, fig.height=5}
output <- compareBs(
  wb.clamp.base,
  wb.clamp.full.fbm,
  celltypeTargets,
  "CLAMPbase",
  "CLAMPfull",
  m = "s"
)

output$plot
```

# Session Information

```{r session-info}
sessionInfo()
```